---
title: "Exploratory Data Analysis"
author: "CJ Piccio and Collin Pucher"
date: '2022-08-04'
output: html_document
---

*This exploratory analysis of America's favorite past-time takes an in-depth look at the history of Negro League play and the statistics they generated.*

```{r include = FALSE}
library(tidyverse)
library(plotly)
library(RColorBrewer)
```

#### Our Data

The data we are using is from Baseball Reference, though it was compiled by Seamheads from newspapers and other primary sources of the time. It is important to note that the data set is not complete by any means, as box scores and record keeping were not as comprehensive in the Negro Leagues as they were for the American and National Leagues. In addition, the Negro Leagues played many exhibition, barnstorming, and other types of games that are not included in the data set. 

Our data includes all Negro League pitchers and position players from 1920 to 1948 (the period they have been designated major leagues by MLB) in the Baseball Reference database, with the aforementioned caveats. Our data set includes all hitters with at least 100 plate appearances (PA), and all pitchers who appeared in at least 10 games. The data set only includes the career stats each player put up while in the Negro Leagues (for instance, Jackie Robinson's statistics with the Dodgers are not included in the set).

#### Summary Information

The median number of plate appearances in our dataset was 374, with the mean being 672.2. The median and mean on-base percentage plus slugging percentage (OPS) was .655 (more on this statistic later). For pitchers, the median earned run average (ERA) was 4.50, with the mean ERA being 4.74. The median fielding independent pitching (FIP) was 2.80, with the mean FIP being 2.77.

```{r echo = FALSE}
all_1920_48_qb <- read.csv("data/NegroLeagueBatters1920to1948.csv")
all_1920_48_qp <- read.csv("data/NegroLeaguePitchers1920to1948.csv")

bat_summary <- summary(all_1920_48_qb)
pitcher_summary <- summary(all_1920_48_qp)

bat_summary
pitcher_summary
```

#### "He gets on-base."

One of the greatest baseball managers of all time, Earl Weaver of the Baltimore Orioles, famously said "your most precious possessions on offense are your 27 outs." On-base percentage (OBP) is a useful statistic because it measures a hitter's ability to avoid an out. It is superior to earlier statistics like batting average because it factors in a batter's ability to draw walks.

$OBP = \frac{H+BB+HBP}{AB+BB+HBP+SF}$

OBP was made famous by Michael Lewis's 2003 book Moneyball (later adapted into a movie), which chronicled the 2002 Oakland Athletics and their data-driven quest to build a competitive roster with a much lower payroll than other MLB teams. In a [memorable scene from the movie](https://www.youtube.com/watch?v=PlKDQqKh03Y), Brad Pitt, playing general manager Billy Beane, describes why he pursues certain players to the older, less data-driven scouts who don't understand. Over their objections to superficial factors like weight or off-field issues, Beane simply states, "He gets on-base."

Here is an interactive visual of OBP on the X-axis and slugging percentage (SLG) (essentially a measure of power) on the Y-axis for Negro Leaguers from 1920 to 1948 with at least 100 plate appearances.

```{r echo = FALSE}
all_1920_48_qb <- read.csv("data/NegroLeagueBatters1920to1948.csv")
all_1920_48_qb <- arrange(all_1920_48_qb, PA)

all_ops_plot_2048 <- ggplot(data = all_1920_48_qb, aes(OBP, SLG, col = PA, label = Player)) + 
                   geom_point(aes(size = PA)) + labs(x = "On-base percentage", y = "Slugging percentage", 
                                       title = "1920 to 1948 Negro League batters, > 100 PA") 
all_ops_plotly_2048 <- ggplotly(all_ops_plot_2048)
all_ops_plotly_2048
```

#### Radar Charts
